---
output: html_document
---
title: "Storm_Assignment"
author: "ISSA FALL"
date:  november,11 2015
output: html_document:
         keep_md:true
         
##                         WEATHER AND ITS RELATED CONSEQUENCES

#INTRODUCTION
```
Across the United States,severe weather events are the headline every year.
Exploring the NOAA Storm Database from 1950 to november 2011,allow me to discover the most harmfull
weather events and their related economic consequences.
```
# DATA PROCESSING
```{r loading ,echo=TRUE}
dir("C:/Users/ndeye amy fall/Desktop/RStudio/repdata_data_StormData.csv.bz2")
```
##loading
```{r Processing,echo=TRUE}
con<-file("repdata_data_StormData.csv.bz2","rt")
storm<-read.csv(con)
str(storm)
summary(storm)
evtype<-storm$EVTYPE
head(evtype)
```
##Subset of variables that are relevant to health or economy and explore some of them
```{r subset,echo=TRUE}
library(dplyr)
library(plyr)
library(ggplot2)
storm2<-storm[, c(7,8,23,24,25,26,27,28)]
##set.seed(900000)
##trainIndicator=rbinom(902297,size = 1,prob = 0.5)
##table(trainIndicator)
##trainStorm=storm[trainIndicator == 1,]
##testStorm=storm[trainIndicator == 0,]
##plot(log10(trainStorm[,c(23,24,25,27)]+1))
##hCluster = hclust(dist(t(log10(trainStorm[,c(23,24,25,27)]+1))))
```
##RESULTS
```{r results,echo=TRUE}
```
##20 most harmfull events with respect to population health
```{r harmfull events,echo=TRUE}
fatalities<-aggregate(FATALITIES ~ EVTYPE, storm, FUN = sum)
fatalities<-head(fatalities[order(fatalities$FATALITIES,na.last = TRUE,decreasing = TRUE),], 20)
fatalities
injuries<-aggregate(INJURIES ~ EVTYPE, storm, FUN = sum)
injuries<-head(injuries[order(injuries$INJURIES,na.last = TRUE,decreasing = TRUE),], 20)
injuries
```
##Total number of fatalities and injuries from 1950 to november 2011
```{r fatalities and injuries}
fatalities1<-summarize(group_by(storm,EVTYPE), FATALITIES =sum(FATALITIES,na.rm = TRUE))
fatalities1
injures1<-summarize(group_by(storm,EVTYPE), INJURIES =sum(INJURIES,na.rm = TRUE))
injures1
```
##20 events that have the greatest economic consequences
```{r economic consequences,echo=TRUE}
propdmg<-aggregate(PROPDMG ~ EVTYPE, storm, FUN = sum)
head(propdmg[order(propdmg$PROPDMG,na.last = TRUE,decreasing = TRUE), ],20)
cropdmg<-aggregate(CROPDMG ~ EVTYPE, storm, FUN = sum)
head(cropdmg[order(cropdmg$CROPDMG,na.last = TRUE,decreasing = TRUE), ], 20)
```
##Total number of properties and quantities of crops domaged from 1950 to november 2011
```{r total expenses,echo=TRUE}
propdmg1<-summarize(group_by(storm,EVTYPE), PROPDMG =sum(PROPDMG,na.rm = TRUE))
propdmg1
cropdmg1<-summarize(group_by(storm,EVTYPE), CROPDMG =sum(CROPDMG,na.rm = TRUE))
cropdmg1
```
##Tables
```{r tables,echo=TRUE}
table(storm$PROPDMGEXP)
table(storm$CROPDMGEXP)
```
##Clean up EVTYPE variable and convert all to upper case
```{r clean up,echo=TRUE}
##storm2$EVTYPE<-gsub("\\S+"," ",storm2$EVTYPE)
##storm2$EVTYPE<-gsub("^\\S+|\\S+$"," ",storm2$EVTYPE)
##storm2$EVTYPE<-toupper(storm2$EVTYPE)
##storm2$EVTYPE<-as.factor(storm2$EVTYPE)
```
##Using the above information for further clean up of EVTYPE
```{r further clean up,echo=TRUE}
storm$EVTYPE<-as.factor(storm$EVTYPE)
storm$PROPDMG[storm$PROPDMGEXP == "H"]<-storm$PROPDMG[storm$PROPDMGEXP == "H"]*100
storm$PROPDMG[storm$PROPDMGEXP == "K"]<-storm$PROPDMG[storm$PROPDMGEXP == "K"]*1000
storm$PROPDMG[storm$PROPDMGEXP == "M"]<-storm$PROPDMG[storm$PROPDMGEXP == "M"]*1000000
storm$PROPDMG[storm$PROPDMGEXP == "B"]<-storm$PROPDMG[storm$PROPDMGEXP == "B"]*1000000000

storm$CROPDMG[storm$CROPDMGEXP == "H"]<-storm$CROPDMG[storm$CROPDMGEXP == "H"]*100
storm$CROPDMG[storm$CROPDMGEXP == "K"]<-storm$CROPDMG[storm$CROPDMGEXP == "K"]*1000
storm$CROPDMG[storm$CROPDMGEXP == "M"]<-storm$CROPDMG[storm$CROPDMGEXP == "M"]*1000000
storm$CROPDMG[storm$CROPDMGEXP == "B"]<-storm$CROPDMG[storm$CROPDMGEXP == "B"]*1000000000
```
##Economic impacts on property and crop domages expenses(H=100,K=1000,M=10^6,B=10^9)
```{r economic impacts,echo=TRUE}

ds<-storm[, c("EVTYPE","FATALITIES","INJURIES","PROPDMG","CROPDMG","PROPDMGEXP","CROPDMGEXP")]
ds2<-!is.na(ds$PROPDMG)
ds<-ds[ds2, ]
head(ds)
dmgByevtype<-aggregate(PROPDMG~EVTYPE,data = ds,sum)
topdmgByevtype<-order(dmgByevtype$PROPDMG,decreasing = TRUE)[1:20]
format(dmgByevtype[topdmgByevtype,c(1,2)],big.mark = ",",scientific = FALSE)
barplot(dmgByevtype[topdmgByevtype, 2],xlab = "Event type",ylab = "Total properties domaged by type",cex.lab=1.3,names.arg = dmgByevtype$EVTYPE[topdmgByevtype])
```
##Amount of dollar domages
```{r dollar domages,echo=TRUE}
##propdmgexp<-aggregate(PROPDMGEXP~EVTYPE,ds,FUN = sum)
##head(propdmgexp[order(propdmgexp$PROPDMGEXP,na.last = TRUE,decreasing = TRUE),], 20)
##cropdmgexp<-aggregate(CROPDMGEXP~EVTYPE,ds,FUN = sum)
##head(cropdmgexp[order(cropdmgexp$CROPDMGEXP,na.last = TRUE,decreasing = TRUE),], 20)
```
##Plot of 5 most harmfull Weather events to population health(INJURIES)
```{r plot,echo=TRUE}
ggplot(injuries[1:5, ],aes(EVTYPE,INJURIES))+geom_bar(fill="red",stat="identity")+ggtitle("5  Most harmfull Weather events(injuries)")+xlab("type of weather events")+ylab("INJURIES")+theme(plot.title=element_text(lineheight = .8,face = "bold"))
```


##Plot of 5 most harmfull weather events to population health(FATALITIES)
```{r most harmfull weather,echo=TRUE}
ggplot(fatalities[1:5, ], aes(EVTYPE,FATALITIES)) + geom_bar(fill="red",stat = "identity")+ggtitle("5 Most harmfull weather events(fatalities)")+xlab("Type of weather events")+ylab("FATALITIES")+theme(plot.title=element_text(lineheight = .8,face = "bold"))
```




